---
title: "Naive Bayes Classifier"
author: "Quan Nguyen"
date: "June 29, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
# Loading libraries
library(tidyverse)
```

# Data simulation
## Step 1 - Collection and Exploring the Data
Generate 100 cases, using two attributes x1 and x2 using Gaussian distribution

```{r}
num_attrib <- 2
N <- 100

# Seed
set.seed(2017)
x <- matrix(rnorm(N * num_attrib), ncol = num_attrib)
colnames(x) <- c("x1", "x2")

# Set class variable
y <- as.numeric((x[, 1] ^ 2 + x[, 2] ^ 2) > 2.3)
```

Inspecting the newly create data and create data frame
```{r}
head(x)
head(y)

#Create data frame
y <- as.factor(y)
data <- as.data.frame(cbind(y, x))
```

Visualize data
```{r}
library(ggplot2)
ggplot(data, aes(x = x1)) +
  geom_density(color = "red")
ggplot(data, aes(x = x2)) +
  geom_density(color = "green")
ggplot(data, aes(x = y)) +
  geom_bar(aes(color = y, fill = y))
```

## Step 2 - Preparing the Data

We're going to select  at random without replacement and put them in **train** set. The rest of the data (3 observation) is going to **test** set.

```{r}
set.seed(2016)

# Without replacement
train <- sample(1:N, 70, FALSE)
```

## Step 3 - Train model using **train** set
```{r}
library(e1071)
fit <- naiveBayes(x[train, ], y[train])
```

## Step 4 - Evaluate model performance
```{r}
# Probability
predc_probs <- predict(fit, data[train, -1], type = "raw")
head(predc_probs)
```

Viewing the predicted class label

```{r}
pred <- predict(fit, data[train, -1], type = "class")
head(pred)
```

Confusion matrix
```{r}
y_train <- y[train]
table(pred, y_train)
```


## Step 5 - How well doest it perform on **test** set
```{r}
pred_test <- predict(fit, data[-train, -1], type = "class")
y_test <- y[-train]
table(pred_test, y_test)
```


# Identifying Radar signals
## Step 1 - Collecting and Exploring the Data
```{r}
data("ionosphere", package = "evclass")
str(ionosphere)
```












