---
title: "ml-easy-decision3"
author: "Quan Nguyen"
date: "June 25, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Collection and Exploring the Data
Loading data from package **mclust**

```{r}
data("banknote", package = "mclust")
# You gonna see genuine banknote at the top
head(banknote)
# Counterfeit banknote at the bottom
tail(banknote)
```
Check the distribution of the *Status* variable
```{r}
table(banknote$Status)
```


View the correlation of all the variable now

```{r corrplot, echo = FALSE}
library(corrplot)
corr_banknote <- cor(banknote[-1])
corrplot(corr_banknote, method = "square")
```

## Step 2 - Preparing the Data

We're going to select 150 at random without replacement and put them in **train** set. The rest of the data (50 observation) is going to **test** set.

```{r}
set.seed(2018)
N <- nrow(banknote)
train <- sample(1:N, 150, FALSE)

# Note that train store only the row numbers not the actual obs
head(train)
```

## Step 3 - Train model using **train** set
```{r}
library(C50)
fitc <- C5.0(Status ~., data = banknote[train, ])
```
Now see the result 
```{r train decision tree, echo = FALSE}
plot(fitc)
```

You can see verbal rule:
```{r}
fitc_rules <- C5.0(Status ~., data = banknote[train, ], rules = TRUE)
summary(fitc_rules)
```

## Step 4 - Evaluate model performance
```{r}
# Get the prediction result
predc_train <- predict(fitc, newdata = banknote[train, ], type = "class")
head(predc_train)
```
How well the decision tree classify the trainning sample?
```{r}
table(banknote$Status[train], predc_train, dnn = c("Observed Class", "Predicted class"))
```
How well doest it perform on **test** set
```{r}
predc <- predict(fitc, newdata = banknote[-train, ], type = "class")
table(banknote$Status[-train], predc, dnn = c("Observed Class", "Predicted Class"))
```
## Step 5 - Improving Model Performance
```{r}
# Tree::tree : Deviance or Gini metric
library(tree)
fit <- tree(Status ~., data = banknote[train, ], split = "deviance")
# View the decision tree
plot(fit)
text(fit) #Display text on decision tree
```
Take a look at train set performance
```{r}
summary(fit)
```

Test set performance
```{r}
pred <- predict(fit, newdata = banknote[-train,])
tail(pred, 5)
pred.class <- colnames(pred)[max.col(pred, ties.method = c("random"))]
# Take a look to see if they match the obs listed in tail function
tail(pred.class, 5)
```
Use confustion matrix now
```{r}
table(banknote$Status[-train], pred.class, dnn = c("Observed Class", "Predicted Class"))
```
Congratulations! That was 100% correct of prediction !