---
title: "K nearest neighbors"
author: "Quan Nguyen"
date: "June 27, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Collection and Exploring the Data
Loading data from package **rebmix**

```{r}
data("wine", package = "rebmix")
# Take a look at the data
str(wine)
head(wine)

# Well, Cultivar is a factor
```
How many Cultivar are there?

```{r}
library(ggplot2)
ggplot(wine, aes(x = Cultivar)) +
  geom_bar(aes (color= Cultivar, fill = Cultivar), stat = "count")
```


View the correlation of all the variable now

```{r corrplot, echo = FALSE}
library(corrplot)
corr_wine <- cor(wine[-14]) #Remove the Cultivar column
corrplot(corr_wine, method = "square")
```

## Step 2 - Preparing the Data
Display raw features by boxplot

```{r}
boxplot(wine)
```

The performance of KNN depends critically on the distance metric. For this reason, you should typically normalize or scale the features so that they are approximately in the same range
```{r}
data_sample <- wine[, 1:13]
data_sample <- scale(data_sample)
boxplot(data_sample)
```

Make **train** and **test** dataset
```{r}
set.seed(2016)
n = nrow(data_sample)
train <- sample(1:n, 89, replace = FALSE)
head(train)
```

## Step 3 - Train model using **train** set
```{r}
require(knnGarden)
fit1 <- knnVCN(data_sample[train,],
               wine$Cultivar[train],
               data_sample[-train,],
               K = 2,
               method = "canberra")

```

## Step 4 - Evaluate model performance
```{r}
# Confusion matrix
table(fit1$TstXIBelong, wine$Cultivar[-train])
```

## Step 5 - Improving Model Performance
Try the Euclidean distance
```{r}
fit2 <- knnVCN(data_sample[train, ],
               wine$Cultivar[train],
               data_sample[-train,],
               K = 2,
               method = "euclidean")
# Confusion matrix
table(fit2$TstXIBelong, wine$Cultivar[-train])
```
Changing K
```{r}
fit3 <- knnVCN(data_sample[train, ],
               wine$Cultivar[train],
               data_sample[-train,],
               K = 3,
               method = "euclidean")
table(fit3$TstXIBelong, wine$Cultivar[-train])
```

Remember the value of K and the selected distance metric can have a significant impact on overall performance of the KNN algorithm