---
title: "Visualizing Tennis Grand Slam Winners Performances"
author: "Giorgio Garziano"
date: "May 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Data visualization of sports historical results is one of the means by which champions strengths and weaknesses comparison can be outlined. In this tutorial, we show what plots flavors may help in champions performances comparison, timeline visualization, player-to-player and player-to-tournament relationships. We are going to use the Tennis Grand Slam Tournaments results as outlined by the ESP site at: [ESPN site tennis history table](http://www.espn.com/tennis/history) and which has been made available as tab-delimited file at the following link: [tennis-grand-slam-winners](https://datascienceplus.com/wp-content/uploads/2017/04/tennis-grand-slam-winners.txt)

Our analysis shall involve basic dataset manipulation as well. Overall, we are going to take advantage of the following R packages:

`library(ggplot2)` #for barplots  
`library(gplots)` #for heatmaps  
`library(RColorBrewer)` #for palettes  
`library(dplyr)` #for dataset manipulation  
`library(knitr)` #for neaty dataset printing  
`library(timelineS)` #for timeline plot  
`library(circlize)` #for chord-diagrams  
`library(fmsb)` #for radar plots  

## Analysis
Loading R libraries and importing the Tennis Grand Slam Winners dataset.
```{r}
library_toload <-
  c(
  "dplyr",
  "knitr",
  "ggplot2",
  "gplots",
  "RColorBrewer",
  "timelineS",
  "circlize",
  "fmsb"
  )
  invisible(lapply(library_toload, function(x) {
  suppressPackageStartupMessages(library(x, character.only = TRUE))
  }))
  
  
  url_file <- "https://datascienceplus.com/wp-content/uploads/2017/04/tennis-grand-slam-winners.txt"
  slam_win <-  read.delim(url(url_file), sep = "\t", stringsAsFactors = FALSE)
  kable(head(slam_win, 20))
```
A minor fix to the tournaments data column is needed to have same naming for the Australian Open tournament.

```{r}
slam_win[grep("Australian Open", slam_win$TOURNAMENT), "TOURNAMENT"] = "Australian Open"
```
## Barplot
Grouping by winner and summarising the number of wins are preliminary steps in order to compute a table where each champion name is associated to his own number of Tennis Grand Slam wins.
```{r}
slam_top_chart <- slam_win %>%
  group_by(WINNER) %>%
  summarise(NUM_WINS = n()) %>%
  arrange(desc(NUM_WINS))

kable(head(slam_top_chart, 40))
```

To graphically introduce such table, a barplot reporting winners ordered by their number of wins is suggestable. In that way, we can evaluate the players performance from the absolute perspective and relative to each other. Our barplot involves the champions who won at least four Tennis Grand Slam tournaments.

```{r}
slam_top_chart$WINNER <-
  factor(slam_top_chart$WINNER, levels = slam_top_chart$WINNER[order(slam_top_chart$NUM_WINS)])
  
top_winners_gt4 <-  slam_top_chart %>%
                    filter(NUM_WINS >= 4)
the_colours <-  c(
  "#FF4000FF",
  "#FF8000FF",
  "#FFFF00FF",
  "#80FF00FF",
  "#00FF00FF",
  "#00FF80FF",
  "#00FFFFFF",
  "#0080FFFF",
  "#FF00FFFF",
  "#000000FF",
  "#0000FFFF"
  )

ggplot(data = top_winners_gt4, aes(x = WINNER, y = NUM_WINS, fill = NUM_WINS)) +
  geom_bar(stat = 'identity') + coord_flip() + guides(fill = FALSE) +
  scale_fill_gradientn(colours = the_colours)

```

Further, to facilitate comparison of champions’ performances for each specific tournament, the grouping by TOURNAMENT and WINNER is necessary.
```{r}
slam_top_chart_by_trn <-  slam_win %>%
  filter(WINNER %in% top_winners_gt4$WINNER) %>%
  group_by(TOURNAMENT, WINNER) %>%
  summarise(NUM_WINS = n()) %>%
  arrange(desc(NUM_WINS))
  
slam_top_chart_by_trn$NUM_WINS <- factor(slam_top_chart_by_trn$NUM_WINS)

kable(head(slam_top_chart_by_trn, 10))
```

To obtain the required barplot flavor, we take specifically advantage of the discrete scale for y axis and the facet_grid based on the TOURNAMENT field.

```{r}
ggplot(data = slam_top_chart_by_trn, aes(x = WINNER, y = NUM_WINS, fill = NUM_WINS)) +
  geom_bar(stat = 'identity') + coord_flip() + guides(fill = FALSE) + scale_y_discrete() +
  facet_grid(. ~ TOURNAMENT)
```

HeatmapTo highlight how many times champions met each other on Grand Slam tournament finals, we may take advantage of an heatmap. The heatmap rows and columns entries are populated with champions names and each cell reports how many times such players met in the tournament final. A gradient colouring is used for better distinguishing the number of wins. We are going to take advantage of the heatmap.2() function made available by the gplots package which provides some more options with respect heatmap(). We consider the fifty most recent sport events.
```{r}
tl_rec <- 1:50
winner_runnerup <- slam_win[tl_rec, c("WINNER", "RUNNER.UP")]
winner_runnerup_names <- unique(c(winner_runnerup[,1], winner_runnerup[,2]))

match_matrix <- matrix(0, nrow=length(winner_runnerup_names),
                       ncol=length(winner_runnerup_names))
colnames(match_matrix) <- winner_runnerup_names
rownames(match_matrix) <- winner_runnerup_names

for(i in 1:nrow(winner_runnerup)) {
  winner <- winner_runnerup[i, "WINNER"]
  runner_up <- winner_runnerup[i, "RUNNER.UP"]
  r <- which(rownames(match_matrix) == winner)
  c <- which(colnames(match_matrix) == runner_up)

  match_matrix[r,c] <- match_matrix[r,c] + 1
  match_matrix[c,r] <- match_matrix[c,r] + 1
}

diag(match_matrix) <- NA
```

That matrix reporting the counting of players’ pairs finals can be used as heatmap.2 function input.
```{r}
my_palette <- colorRampPalette(c("green", "yellow", "red"))(n = 299)
col_breaks = c(seq(0, 0.99, length=100),  # for green
               seq(1, 5, length=100),     # for yellow
               seq(5.01, 10, length=100)) # for red
heatmap.2(match_matrix,
          cellnote = match_matrix,  # same data set for cell labels
          main = "Tennis Grand Slam Champions - Finals Match Heatmap", # heat map title
          notecol = "black",      # change font color of cell labels to black
          density.info = "none",  # turns off density plot inside color legend
          trace = "none",         # turns off trace lines inside the heat map
          margins = c(12,9),     # widens margins around plot
          col= my_palette,       # use on color palette defined earlier
          breaks = col_breaks,    # enable color transition at specified limits
          dendrogram = "none",     # only draw a row dendrogram
          Colv = "NA")

```

## Dendrogram plot
In case we would like to group champions based on some specific similarity metric and show the result, we can take advantage of the dendrogram plot. So, considering the first twenty champions of our top list as determined in previous steps, we get the following table:

```{r}
ch_n <- 1:20
kable(slam_top_chart[ch_n, ])
```

We specify an euclidean metrics distance for clustering based on champions’ wins. Groups are determined based on wins difference equal to two, (see h_value parameter).

```{r}
wins <- slam_top_chart[ch_n,-1]
d_wins <- dist(wins, method = "euclidean")
hclust_fit <- hclust(d_wins)
h_value <- 2
groups <- cutree(hclust_fit, h = h_value)
plot(hclust_fit, labels = slam_top_chart$WINNER[ch_n], main = "Champions Dendrogram")
rect.hclust(hclust_fit, h = h_value, border = "blue") 

```


## Timeline plot
The timeline plot may be the right choice to highlight how the wins sequence happened. Our timeline plot shows winners’ names together with tournaments and calendar dates. All that considering the twenty most recent results. The year column data is enhanced to match the date format.
```{r}
year_to_date_trnm <- function(the_year, the_trnm) {
  the_date <- NULL
  if (the_trnm == "Australian Open") {
    the_date <- (paste(the_year, "-01-31", sep=""))
  } else if (the_trnm == "French Open") {
    the_date <- (paste(the_year, "-06-15", sep=""))
  } else if (the_trnm == "Wimbledon") {
    the_date <- (paste(the_year, "-07-15", sep=""))
  } else if (the_trnm == "U.S. Open") {
    the_date <- (paste(the_year, "-09-07", sep=""))
  }
  the_date
}

slam_win$YEAR_DATE <- as.Date(mapply(year_to_date_trnm, slam_win$YEAR, slam_win$TOURNAMENT))
tl_rec <- 1:20
timelineS(slam_win[tl_rec, c("WINNER", "YEAR_DATE")], line.color = "red", scale.font = 3,
          scale = "month", scale.format = "%Y", label.cex = 0.7, buffer.days = 100,
          labels = paste(slam_win[tl_rec, "WINNER"], slam_win[tl_rec, "TOURNAMENT"]))

```

## Chord Diagram
Suppose you want to highlight for relationship between champions and tournaments wins while having an idea of the reciprocal strength in that. The chord diagram may be a good solution for such purpose. Herein below, we filter
out a dataset where champions with more than ten Grand Slam tournaments wins are encompassed.
```{r}
top_winners_gt10 = slam_top_chart %>% filter(NUM_WINS > 10)
kable(head(top_winners_gt10))
```

We then take advantage of an inner join between the Grand Slam dataset we started with and the list above. Further, the grouping by tournament and winner and the computation of total wins per tournament gives the following table.
```{r}
slam_win_cnt <-  inner_join(slam_win, top_winners_gt10) %>%
  select(TOURNAMENT, WINNER) %>%
  group_by(WINNER, TOURNAMENT) %>%
  summarise(NUM_WINS = n()) %>%
  arrange(TOURNAMENT, desc(NUM_WINS))
kable(slam_win_cnt)
```

The latter dataset is the input to the chordDiagram function.

```{r}
chordDiagram(slam_win_cnt)
```

## Radar plot
In case we want to highlight champions’ strengths and weaknesses in each specific Grand Slam tournament, a radar plot may be a good choice. Herein below, the radar plots associated to well known tennis champions are shown.
```{r}
champion_radar_plot <- function(df, champion_name) {
  # Filter data for the champion
  slam_win_cnt_chp = df %>% filter(WINNER == champion_name)
  chp_num_wins <- slam_win_cnt_chp$NUM_WINS
  l <- length(chp_num_wins)
  max_v <- 10 # choosing the same maximum value for all champions
  chp_df <-  data.frame(rbind(max = rep(max_v, l), min = rep(0, l), chp_num_wins))
  colnames(chp_df) <- slam_win_cnt_chp$TOURNAMENT
  seg_n <- max_v
  radarchart(
    chp_df,
    axistype = 1,
    caxislabels = seq(0, max_v, 1),
    seg = seg_n,
    centerzero = TRUE,
    pcol = rgb(0.2, 0.5, 0.5, 0.9) ,
    pfcol = rgb(0.2, 0.5, 0.5, 0.3),
    plwd = 1,
    cglcol = "grey",
    cglty = 1,
    axislabcol = "blue",
    vlcex = 0.8,
    calcex = 0.7,
    title = champion_name
    )
  
}

champion_radar_plot(slam_win_cnt, "Roger Federer")
champion_radar_plot(slam_win_cnt, "Rafael Nadal")
champion_radar_plot(slam_win_cnt, "Novak Djokovic")
```

## Conclusions
In this tutorial, we showed different flavors of plots capable to capture many insights of our Tennis Grand Slam tournaments dataset. By means of basic preliminary dataset manipulation and powerful plots, we obtained effective visualizations that can drive the audience attention right to the messages we want to deliver to.
If you have any questions, please feel free to comment below.


