---
title: "Introduction to ggplot2"
author: "Hadley Wickham"
date: "November 11, 2016"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring the mpg data set
```{r warning = FALSE, message = FALSE}
library(ggplot2)
head(mpg)
str(mpg)
summary(mpg)
```

## Now let's try first plot
```{r}
qplot(displ, hwy, data = mpg)
```

**Differentiate class by colour**
```{r}
qplot(displ, hwy, colour = class, data = mpg)
```

## Experiment with different aesthetics: colours, size, shape
### Set colour to a continuous variable
```{r}
qplot(displ, hwy, colour = cty, data = mpg)
```

### Using shape
```{r error=TRUE}
qplot(displ, hwy, shape = class, data = mpg)
qplot(displ, hwy, shape = cyl, data = mpg)
```

### Using size
```{r}
qplot(displ, hwy, size = cyl, data = mpg)
```

### Combination
```{r}
qplot(displ, hwy, shape = drv, colour = drv, data = mpg)
```

## Faceting
```{r}
qplot(displ, hwy, data = mpg) + facet_grid(. ~ cyl)
qplot(displ, hwy, data = mpg) + facet_grid(drv ~ .)
qplot(displ, hwy, data = mpg) + facet_grid(drv ~ cyl)
qplot(displ, hwy, data = mpg) + facet_wrap(~ class)
```

## Overplotting problem
```{r}
qplot(cty, hwy, data = mpg)
qplot(cty, hwy, data= mpg, geom = "jitter")
```

## Using reorder and layers

```{r}
qplot(class, hwy, data= mpg)
qplot(class, hwy, data= mpg, geom = "jitter")
qplot(class, hwy, data= mpg, geom = "violin")
qplot(class, hwy, data= mpg, geom = "boxplot")
qplot(reorder(class, hwy), hwy, data= mpg, geom = "boxplot")

#reorder by number of observation
qplot(reorder(class, hwy, length), hwy, data= mpg, geom = "boxplot")
qplot(reorder(class, hwy, median), hwy, data= mpg, geom = "boxplot")
qplot(reorder(class, hwy, median), hwy, data= mpg, geom = c("jitter","boxplot"))

#put jitter layer in front of boxplot
qplot(reorder(class, hwy, median), hwy, data= mpg, geom = c("boxplot","jitter"), colour = drv)
```
## Exploring the diamonds datasetstr(diamonds)
```{r}
summary(diamonds)
head(diamonds)
tail(diamonds)
?diamonds
```

With only one variable, ggplot2 guesses that you want a barchart or histogram

```{r}
qplot(cut, data = diamonds)
qplot(carat, data = diamonds)
# Always testing with binwidth for best visualization
qplot(carat, data = diamonds, binwidth = 1)
qplot(carat, data = diamonds, binwidth = 0.1)
qplot(carat, data = diamonds, binwidth = 0.01)
# This will tell you the resolution of the variable
resolution(diamonds$carat)
last_plot() + xlim(0,3)
```

Explore the depth by cut
```{r}
qplot(depth, data = diamonds, binwidth = 0.2)
qplot(depth, data = diamonds, binwidth = 0.2, fill = cut) + xlim(55,70)
qplot(depth, data = diamonds, binwidth = 0.2) + facet_wrap(~ cut)
# If you want to see how much binwidth is OK then take a look at its range
range(diamonds$price)
# Large distance make comparison hard
qplot(price, data = diamonds, binwidth = 500) + facet_wrap(~ cut)
# Stacked heighs are difficult to compare
qplot(price, data = diamonds, binwidth = 500, fill = cut)
# Much better but still have differing value abundance
qplot(price, data = diamonds, binwidth = 500, geom = "freqpoly", colour = cut)
# Instead of display count. Display density
# .. indicates that variable isn't in original data
qplot(price, ..density.., data = diamonds, binwidth = 500, geom = "freqpoly", colour = cut)
```

To set aesthetics to a particular value, you need to wrap that value in `I()` Otherwise qplot will try to map that to your dataset
```{r}
qplot(carat, price, data = diamonds, colour = "blue")
qplot(carat, price, data = diamonds, colour = I("blue"))
```

Practical application, varying alpha
```{r}
qplot(carat, price, data = diamonds, alpha = I(1/10))
qplot(carat, price, data = diamonds, alpha = I(1/50))
qplot(carat, price, data = diamonds, alpha = I(1/100))
qplot(carat, price, data = diamonds, alpha = I(1/250))

```

There are two ways to add additional geom
```{r}
# A vector of geom name
qplot(price, carat, data = diamonds, geom = c("point","smooth"))
# Add on extra geom
qplot(price, carat, data = diamonds) + geom_smooth()
```

This is how you get help for a specific geom (or go to [ggplot2 documentation](http://docs.ggplot2.org/current))
```{r}
?geom_smooth
```

Now explore the relationship between carat, price, clarity
```{r}
qplot(price, carat, data = diamonds, colour = cut)
qplot(log10(price), log10(carat), data = diamonds, colour = cut)
# Install.packages(hexbin)
qplot(log10(price), log10(carat), data = diamonds, geom = "bin2d", bins = 50) + facet_wrap(~ cut)

qplot(log10(price), log10(carat), data = diamonds, geom = "smooth", colour = cut)

lm(log10(price) ~ log10(carat), data = diamonds)
qplot(log10(price), log10(carat), data = diamonds) + facet_wrap(~ cut) + geom_abline(colour = "red", intercept = 3.6, slope = 1.7)
```

