---
title: "Two groups comparison"
author: "Robert Kabakoff"
date: "November 12, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploring dataset
Load libraries, dataset UScrime is in MASS package
```{r warning = FALSE}
library(MASS)
library(ggplot2)
        
```
Take a look at the data
```{r}
head(UScrime)
str(UScrime)
```

## Independent t-test
Create a factor for So (Southern) and assign it to Region variable in UScrime
```{r}
UScrime$Reg <- factor(UScrime$So,
                      levels = c(0,1),
                      labels = c("Not South", "South"))
```
Now let's see how difference there are in Prob imprisonment in different region
```{r}
qplot(Reg, Prob, data = UScrime, geom = c("boxplot","point"))
```

Formal test to see where it is significant. It turns out YES
```{r}
t.test(Prob ~ Reg, data = UScrime)
```
Because the outcome variable is a proportion, you might try to transform it to normality before carrying out the t-test. Let's do this
```{r}
t.test(log(Prob / (1-Prob)) ~ Reg, data = UScrime)
```
The results are pretty much unchanged

## Dependent t-test
Question: Unemployment rate for youger males U1 (14-24) is greater than for older males U2 (35-39)
```{r}
# Calculate mean and sd from the dataset
sapply(UScrime[c("U1","U2")], function(x)(c(mean = mean(x), sd = sd(x))))
with(UScrime, t.test(U1, U2, paired = TRUE))
```
Younger males have higher rate of unemployment

## Non parametric tests
If the outcome variables are severly skewed or ordinal in nature, you may use non parametric tests
```{r warning = FALSE}
# Incarcenation rate in Southern and non Southern states
with(UScrime, by(Prob, So, median))
# Two independent groups. Non parametric test Wilcoxon rank sum test. Are they different?
wilcox.test(Prob ~ So, data = UScrime)
# Two dependent groups. Calculate median rate of unemployment rate from dataset
sapply(UScrime[c("U1","U1")], median)
# For dependent sample use Wilcoxon signed rank test
with(UScrime, wilcox.test(U1, U2, paired = TRUE))
```

## Comparing more than two groups Kruskal Wallis test
```{r}
# Get region from the original dataset
states <- as.data.frame(cbind(state.region, state.x77))
# Apply the test now
kruskal.test(Illiteracy ~ state.region, data = states)
```